HTW Model

### 2024-02-16 Model/htw_model.qmd
tweaks

Although simple in the abstract, implementations of ABC require researchers to make a number of non-trivial decisions as to i) the discrepancy function between observed and simulated data, ii) whether to compute the discrepancy between trial level data, or a summary statistic of the datasets, iii) the value of the minimum tolerance $\epsilon$. For the present work, we follow the guidelines from published ABC tutorials [@turnerTutorialApproximateBayesian2012; @farrellComputationalModelingCognition2018]. For the test stage, we summarized datasets with mean velocity of each band in the observed dataset as $V_{obs}^{(k)}$ and in the simulated dataset as $V_{sim}^{(k)}$, where $k \in \{1, 2, 3, 4, 5, 6\}$ represents each of the six velocity bands. For computing the discrepancy between datasets in the training stage, we aggregated training trials into three equally sized blocks (separately for each velocity band in the case of the varied group). After obtaining the summary statistics of the simulated and observed datasets, the discrepancy was computed as the mean of the absolute difference between simulated and observed datasets (@eq-discrep-test and @eq-discrep-train).

@@

The final component of our ABC implementation is the determination the appropriate value of $\epsilon$. The setting of $\epsilon$ exerts strong influence on the approximated posterior distribution. Smaller values of $\epsilon$ increase the rejection rate, and improve the fidelity of the approximated posterior, while larger values result in an ABC sampler that reproduces the prior distribution. Because the individual participants in our dataset differed substantially in terms of the noisiness of their data, we employed an adaptive tolerance setting strategy to tailor $\epsilon$ to each individual. The initial value of $\epsilon$ was set to the overall standard deviation of each individuals velocity values. Thus, sampled parameter values that generated simulated data within a standard deviation of the observed data were accepted, while worse performing parameters were rejected. After every 300 samples the tolerance was allowed to increase only if the current acceptance rate of the algorithm was less than 1%. In such cases, the tolerance was shifted towards the average discrepancy of the 5 best samples obtained thus far. To ensure the acceptance rate did not become overly permissive, $\epsilon$ was also allowed to decrease every time a sample was accepted into the posterior.

@@

input_layer <<- output_layer <<-  c(100,350,600,800,1000,1200)
ids2 <- c(1,66,36)
file_name <- "n_iter_200_ntry_300_5354"

@@

### Group level fits

@@

The residuals of model predictions for the testing stage (@fig-htw-resid-pred) show the opposite pattern of fitting method - with models fit only to the test data showing the best performance, followed by models fit to both training and test data, and with models fit only to the training data showing the worst performance (note that y axes are scaled different between plots). Unsurprisingly, the advantage of EXAM is strongest for extrapolation positions (the three smallest bands for both groups - as well as the two highest bands for the Constant group). Although EXAM tends to perform better for both Constant and Varied participants (see also @tbl-htw-modelError), the relative advantage of EXAM is generally larger for the Constant group - a pattern consistent across all three fitting methods. Panel B of @fig-htw-resid-pred directly compares the aggregated observed data to the posterior predictive distributions for the testing stage. Of interest are a) the extent to which the median estimates of the ALM and EXAM posteriors deviate from the observed medians for each velocity band; b) the ability of ALM and EXAM to discriminate between velocity bands; c) the relative performance of models that are constrained by the training data (i.e. the 'fit to train only' and 'fit to both' models) compared to the 'fit to test only' models; and d) the extent to which the variance of the posterior predictive distributions mimics the variance of the observed data.

@@

@@

>>>>

Although simple in the abstract, implementations of ABC require researchers to make a number of non-trivial decisions as to i) the discrepancy function between observed and simulated data, ii) whether to compute the discrepancy between trial level data, or a summary statistic of the datasets, iii) the value of the minimum tolerance $\epsilon$ between simulated and observed data. For the present work, we follow the guidelines from previously published ABC tutorials [@turnerTutorialApproximateBayesian2012; @farrellComputationalModelingCognition2018]. For the test stage, we summarized datasets with mean velocity of each band in the observed dataset as $V_{obs}^{(k)}$ and in the simulated dataset as $V_{sim}^{(k)}$, where $k$ represents each of the six velocity bands. For computing the discrepancy between datasets in the training stage, we aggregated training trials into three equally sized blocks (separately for each velocity band in the case of the varied group). After obtaining the summary statistics of the simulated and observed datasets, the discrepancy was computed as the mean of the absolute difference between simulated and observed datasets (@eq-discrep-test and @eq-discrep-train). For the models fit to both training and testing data, discrepancies were computed for both stages, and then averaged together.

@@

The final component of our ABC implementation is the determination of an appropriate value of $\epsilon$. The setting of $\epsilon$ exerts strong influence on the approximated posterior distribution. Smaller values of $\epsilon$ increase the rejection rate, and improve the fidelity of the approximated posterior, while larger values result in an ABC sampler that simply reproduces the prior distribution. Because the individual participants in our dataset differed substantially in terms of the noisiness of their data, we employed an adaptive tolerance setting strategy to tailor $\epsilon$ to each individual. The initial value of $\epsilon$ was set to the overall standard deviation of each individuals velocity values. Thus, sampled parameter values that generated simulated data within a standard deviation of the observed data were accepted, while worse performing parameters were rejected. After every 300 samples the tolerance was allowed to increase only if the current acceptance rate of the algorithm was less than 1%. In such cases, the tolerance was shifted towards the average discrepancy of the 5 best samples obtained thus far. To ensure the acceptance rate did not become overly permissive, $\epsilon$ was also allowed to decrease every time a sample was accepted into the posterior.

@@

input_layer <<- output_layer <<-  c(100,350,600,800,1000,1200)
ids2 <- c(1,66,36)
file_name <- "n_iter_200_ntry_300_5354"
#file_name <- "n_iter_400_ntry_100_2944"

@@

### Group level aggregations

@@

The residuals of the model predictions for the testing stage (@fig-htw-resid-pred) also show a predictable pattern of fitting method - with models fit only to the test data showing the best performance, followed by models fit to both training and test data, and with models fit only to the training data showing the worst performance (note that y axes are scaled different between plots). Unsurprisingly, the advantage of EXAM is strongest for extrapolation positions (the three smallest bands for both groups - as well as the two highest bands for the Constant group). Although EXAM tends to perform better for both Constant and Varied participants (see also @tbl-htw-modelError), the relative advantage of EXAM is generally larger for the Constant group - a pattern consistent across all three fitting methods. Panel B of @fig-htw-resid-pred directly compares the aggregated observed data to the posterior predictive distributions for the testing stage. Of interest are a) the extent to which the median estimates of the ALM and EXAM posteriors deviate from the observed medians for each velocity band; b) the ability of ALM and EXAM to discriminate between velocity bands; c) the relative performance of models that are constrained by the training data (i.e. the 'fit to train only' and 'fit to both' models) compared to the 'fit to test only' models; and d) the extent to which the variance of the posterior predictive distributions mimics the variance of the observed data.

@@

- ** does EXAM do better for the Constant group because the constant group performs better? Or does training with a single example encourage an exam sort of strategy?

@@

### Accounting for individual patterns
